Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-25 paper/zhang-cook-laa-langrene-menendez.Rmd
some changes based on patricia's comment s

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: |
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object in the optimisation of projection pursuit guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and this workflow can be applied to a broad class of optimisers, to assess their performance. An R package, \CRANpkg{ferrn} [@ferrn], has been created to implement the diagnostics.

@@

The remainder of the paper is organised as follows. 
Section \ref{optim} provides an overview of optimisation methods, specifically line search methods. 
Section \ref{tour} reviews projection pursuit guided tour, defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data structure is defined to capture information during the optimisation, and different diagnostic plots are designed with the data collected.
Section \ref{application} shows applications of how these diagnostic plots can be used to discover interesting aspects of different optimisers and guide improvements to the existing algorithms.
Finally, Section \ref{implementation} describes the R package: \CRANpkg{ferrn} [@ferrn], that implements the visual diagnostics.

@@

The optimisation problem of our interest is on constrained optimization [@bertsekas2014constrained] as defined in the introduction section, and assuming it is not possible to find a solution to the problem in the way of a closed-form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathcal{A}$ space.

A large class of methods uses the gradient information of the objective function, with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the limited complexity of the constraints. As will be shown in the next section, the independent variables in our optimisation problem are the entries of a projection matrix and the computational time required to perform differentiation on a matrix could impede the render of tour animation. Hence, gradient-based methods are not the focus of this paper and considerations will be given to derivative-free methods.

@@

Random search methods [@zabinsky2013stochastic] have a random sampling component as part of their algorithm and have been shown to have the ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sufficient sampling from the entire space would require a long time for convergence to occur, despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have thus been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating machanism. Creeping random search [@White1971-ci] is such a variation, where a candidate point is generated within a neighbourhood of the current point. This makes creeping random search faster to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], introduces a different updating mechanism. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where worse candidates still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @belisle1992convergence; @granville1994simulated] and the global optimum can be attained under mild conditions.

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) in exploratory data analysis. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low-dimensional projections. Guided tour is one variation of a broader class of data visualisation methods, tour, which displays high-dimensional data through a series of animated projections.

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal since theoretical proofs from @diaconis1984asymptotics have shown that projections tend to be normal, as $n$ and $p$ approach infinity under certain conditions. Index functions proposed in the litearture include early indices that can be categorised by measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indices [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

@@

A data structure for diagnosing optimisers in projection pursuit guided tour is first defined. With this data structure, four types of diagnostic plots are presented. In the section, examples will be given to show how these plot can be used to reveal issues in the projection pursuit optimisation.

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: |
  Guided tour [@cook1995grand] searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While much work has been done in proposing  indices in the literature, less has been done on evaluating the performance of the optimisers. In this paper, we propose a data collection object in the optimisation of projection pursuit guided tour and introduce visual diagnostics for examining the performance of the optimisation procedure. These diagnostics and this workflow can be applied to a broad class of optimisers to assess their performance. An R package, \CRANpkg{ferrn} [@ferrn], has been created to implement the diagnostics.

@@

The remainder of the paper is organised as follows. 
Section \ref{optim} provides an overview of optimisation methods, specifically line search methods. 
Section \ref{tour} reviews projection pursuit guided tour, defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data structure is defined to capture information during the optimisation, and different diagnostic plots are designed with the data collected.
Section \ref{application} shows applications of how these diagnostic plots can be used to examine the performance of different optimisers and guide improvements to the existing algorithms.
Finally, Section \ref{implementation} describes the R package: \CRANpkg{ferrn} [@ferrn], that implements the visual diagnostics.

@@

The optimisation problem considered in this paper is on constrained optimization [@bertsekas2014constrained] and assuming it is not possible to find a solution to the problem in the way of a closed-form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathcal{A}$ space.

Gradient-based methods are commonly used to optimise an objective function, with the most notable one being the gradient ascent (descent) method. Although these methods are popular, they rely on the availability of the objective function derivatives. As will be shown in the next section, the independent variables in our optimisation problem are the entries of a projection matrix and the computational time required to perform differentiation on a matrix could impede the render of tour animation. Hence, gradient-based methods are not the focus of this paper and considerations will be given to derivative-free methods.

@@

Random search methods [@zabinsky2013stochastic; @Romeijn2009] have a random sampling component as part of their algorithm and have been shown to have the ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sufficient sampling from the entire space would require a long time for convergence to occur, despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have thus been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating machanism. Creeping random search [@White1971-ci] is such a variation, where a candidate point is generated within a neighbourhood of the current point. This makes creeping random search faster to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], introduces a different updating mechanism. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where worse candidates still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @belisle1992convergence; @granville1994simulated] and the global optimum can be attained under mild regularity conditions.

@@

Projection pursuit guided tour combines two different methods (projection pursuit and guided tour) to explore interesting features in a high dimensional space. Projection pursuit, coined by @friedman1974projection, detects interesting structures (e.g. clustering, outliers and skewness) in multivariate data via low-dimensional projections. Guided tour [@cook2008grand; @cook1995grand] is one variation of a broader class of data visualisation methods, tour [@buja2005computational], which displays high-dimensional data through a series of animated projections.

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal since theoretical proofs from @diaconis1984asymptotics have shown that projections tend to be normal, as $n$ and $p$ approach infinity under certain conditions. There has been plentiful index functions proposed in the literature and here list a few: early indices that can be categorised as measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indices [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

@@

A data structure for diagnosing optimisers in projection pursuit guided tour is first defined. With this data structure, four types of diagnostic plots are presented.

### 2021-03-25 paper/zhang-cook-laa-langrene-menendez.tex
some changes based on patricia's comment s

\abstract{%
Guided tour searches for interesting low-dimensional views of
high-dimensional data via optimising a projection pursuit index
function. The first paper of projection pursuit by
\citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we
implement a data collection object in the optimisation of projection
pursuit guided tour and introduce visual diagnostics based on the data
object collected. These diagnostics and this workflow can be applied to
a broad class of optimisers, to assess their performance. An R package,
\CRANpkg{ferrn} \citep{ferrn}, has been created to implement the
diagnostics.

@@

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to discover interesting aspects of different optimisers and
guide improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \CRANpkg{ferrn}
\citep{ferrn}, that implements the visual diagnostics.

@@

The optimisation problem of our interest is on constrained optimization
\citep{bertsekas2014constrained} as defined in the introduction section,
and assuming it is not possible to find a solution to the problem in the
way of a closed-form. That is, the problem consists of finding the
minimum or maximum of a function \(f \in L^p\) in the constrained
\(\mathcal{A}\) space.

A large class of methods uses the gradient information of the objective
function, with the most notable one being the gradient ascent (descent)
method. Although gradient optimization methods are popular, they rely on
the availability of the objective function derivatives and on the
limited complexity of the constraints. As will be shown in the next
section, the independent variables in our optimisation problem are the
entries of a projection matrix and the computational time required to
perform differentiation on a matrix could impede the render of tour
animation. Hence, gradient-based methods are not the focus of this paper
and considerations will be given to derivative-free methods.

@@

Random search methods \citep{zabinsky2013stochastic} have a random
sampling component as part of their algorithm and have been shown to
have the ability to optimise non-convex and non-smooth functions. The
initial random search algorithm, pure random search
\citep{Brooks1958-zk}, draws candidate points from the entire space
without using any information of the current position and updates the
current position when an improvement on the objective function is made.
As the dimension of the space becomes larger, sufficient sampling from
the entire space would require a long time for convergence to occur,
despite a guaranteed global convergence \citep{spall2005introduction}.
Various algorithms have thus been developed to improve pure random
search by either concentrating on a narrower sampling space or using a
different updating machanism. Creeping random search
\citep{White1971-ci} is such a variation, where a candidate point is
generated within a neighbourhood of the current point. This makes
creeping random search faster to compute, however, global convergence is
not longer guaranteed. On the other hand, simulated annealing
\citep{kirkpatrick1983optimization, bertsimas1993simulated}, introduces
a different updating mechanism. Rather than only updating the current
point when an improvement is made, simulated annealing uses a Metropolis
acceptance criterion, where worse candidates still have a chance to be
accepted. The convergence of simulated annealing algorithms has been
widely researched
\citep{mitra1986convergence, belisle1992convergence, granville1994simulated}
and the global optimum can be attained under mild conditions.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) in exploratory data analysis.
Projection pursuit, coined by \citet{friedman1974projection}, detects
interesting structures (e.g.~clustering, outliers and skewness) in
multivariate data via low-dimensional projections. Guided tour is one
variation of a broader class of data visualisation methods, tour, which
displays high-dimensional data through a series of animated projections.

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. An interesting projection shows structures that
are non-normal since theoretical proofs from
\citet{diaconis1984asymptotics} have shown that projections tend to be
normal, as \(n\) and \(p\) approach infinity under certain conditions.
Index functions proposed in the litearture include early indices that
can be categorised by measuring the \(L^2\) distance between the
projection and a normal distribution: Legendre index
\citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, and natural Hermite index
\citep{cook1993projection}; chi-square index \citep{posse1995projection}
for detecting spiral structure; kurtosis index \citep{Loperfido2020} and
skewness index \citep{Loperfido2018} for detecting outliers in financial
time series; and most recently, scagnostic indices \citep{laa2020using}
for summarising structures in scatterplot matrices based on eight
scagnostic measures.

>>>>

\abstract{%
Guided tour \citep{cook1995grand} searches for interesting
low-dimensional views of high-dimensional data via optimising a
projection pursuit index function. The first paper of projection pursuit
by \citet{friedman1974projection} stated that ``the technique used for
maximising the projection index strongly influences both the statistical
and the computational aspects of the procedure.'' While much work has
been done in proposing indices in the literature, less has been done on
evaluating the performance of the optimisers. In this paper, we propose
a data collection object in the optimisation of projection pursuit
guided tour and introduce visual diagnostics for examining the
performance of the optimisation procedure. These diagnostics and this
workflow can be applied to a broad class of optimisers to assess their
performance. An R package, \CRANpkg{ferrn} \citep{ferrn}, has been
created to implement the diagnostics.

@@

The remainder of the paper is organised as follows. Section \ref{optim}
provides an overview of optimisation methods, specifically line search
methods. Section \ref{tour} reviews projection pursuit guided tour,
defines the optimisation problem and outlines three existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics. A data
structure is defined to capture information during the optimisation, and
different diagnostic plots are designed with the data collected. Section
\ref{application} shows applications of how these diagnostic plots can
be used to examine the performance of different optimisers and guide
improvements to the existing algorithms. Finally, Section
\ref{implementation} describes the R package: \CRANpkg{ferrn}
\citep{ferrn}, that implements the visual diagnostics.

@@

The optimisation problem considered in this paper is on constrained
optimization \citep{bertsekas2014constrained} and assuming it is not
possible to find a solution to the problem in the way of a closed-form.
That is, the problem consists of finding the minimum or maximum of a
function \(f \in L^p\) in the constrained \(\mathcal{A}\) space.

Gradient-based methods are commonly used to optimise an objective
function, with the most notable one being the gradient ascent (descent)
method. Although these methods are popular, they rely on the
availability of the objective function derivatives. As will be shown in
the next section, the independent variables in our optimisation problem
are the entries of a projection matrix and the computational time
required to perform differentiation on a matrix could impede the render
of tour animation. Hence, gradient-based methods are not the focus of
this paper and considerations will be given to derivative-free methods.

@@

Random search methods \citep{zabinsky2013stochastic, Romeijn2009} have a
random sampling component as part of their algorithm and have been shown
to have the ability to optimise non-convex and non-smooth functions. The
initial random search algorithm, pure random search
\citep{Brooks1958-zk}, draws candidate points from the entire space
without using any information of the current position and updates the
current position when an improvement on the objective function is made.
As the dimension of the space becomes larger, sufficient sampling from
the entire space would require a long time for convergence to occur,
despite a guaranteed global convergence \citep{spall2005introduction}.
Various algorithms have thus been developed to improve pure random
search by either concentrating on a narrower sampling space or using a
different updating machanism. Creeping random search
\citep{White1971-ci} is such a variation, where a candidate point is
generated within a neighbourhood of the current point. This makes
creeping random search faster to compute, however, global convergence is
not longer guaranteed. On the other hand, simulated annealing
\citep{kirkpatrick1983optimization, bertsimas1993simulated}, introduces
a different updating mechanism. Rather than only updating the current
point when an improvement is made, simulated annealing uses a Metropolis
acceptance criterion, where worse candidates still have a chance to be
accepted. The convergence of simulated annealing algorithms has been
widely researched
\citep{mitra1986convergence, belisle1992convergence, granville1994simulated}
and the global optimum can be attained under mild regularity conditions.

@@

Projection pursuit guided tour combines two different methods
(projection pursuit and guided tour) to explore interesting features in
a high dimensional space. Projection pursuit, coined by
\citet{friedman1974projection}, detects interesting structures
(e.g.~clustering, outliers and skewness) in multivariate data via
low-dimensional projections. Guided tour
\citep{cook2008grand, cook1995grand} is one variation of a broader class
of data visualisation methods, tour \citep{buja2005computational}, which
displays high-dimensional data through a series of animated projections.

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. An interesting projection shows structures that
are non-normal since theoretical proofs from
\citet{diaconis1984asymptotics} have shown that projections tend to be
normal, as \(n\) and \(p\) approach infinity under certain conditions.
There has been plentiful index functions proposed in the literature and
here list a few: early indices that can be categorised as measuring the
\(L^2\) distance between the projection and a normal distribution:
Legendre index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial}, and natural Hermite index
\citep{cook1993projection}; chi-square index \citep{posse1995projection}
for detecting spiral structure; kurtosis index \citep{Loperfido2020} and
skewness index \citep{Loperfido2018} for detecting outliers in financial
time series; and most recently, scagnostic indices \citep{laa2020using}
for summarising structures in scatterplot matrices based on eight
scagnostic measures.
